From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction

### 2015-09-21 paper.tex
Re-organization, formatting, starting to work on 4.2

\section{Experimental Validation}

@@

Six children (ages 5 to 6, 3 boys, 3 girls) were enrolled for this study.
The study took place at school, in a separate room (the computer
lab). The participants were chosen by the teacher, and would come one after the
other for a face-to-face interaction with a {\sc Nao} robot (average duration:
19.6 min, $SD:1.58$).

The experiment is organized in rounds of writing: during a typical round, the
child requests the robot to write something (a single letter, a number or a full
word), and pushes a tactile tablet (equipped with a custom application that we
developed) towards the robot; the robot writes on the tablet by mimicking the
writing gesture (but without actually physically touching the tablet); the child
then pull back the tablet, corrects the robot's attempt by writing him/herself
on top or next to the robot's writing, and ``send'' his/her demonstration to the
robot by pressing a small button on the tablet. The robot learns from this
demonstration and tries again. The child continues the turn-taking until he/she
decides to train the robot on another word. In total, the children performed in
average 12.16 ($SD:2.61$) rounds of writing.

@@

The experiment was carried with an Aldebaran's {\sc Nao} robot, using ROS as a
middleware to build the attention estimation pipeline
(Figure~\ref{fig:system}). Head pose estimation builds on the {\tt dlib} and
OpenCV libraries, while pose transformations are handled by the {\sc tf}
library. The implementation is open-source and available at
\url{https://github.com/chili-epfl/attention-tracker}.

The implementation of the hand-writing activity itself has been previously
presented in~\cite{Hood:2015}.

@@

We performed a {\it post-hoc} coding the video records to manually of the
interaction were (note that the focus of attention was separately hand-coded,
{\it post-hoc}, from the video recording of the interaction).

\subsubsection{From Face Detection to Focus of Attention} \label{fromFaceTo}

Our results for a peer-to-peer work scenarios using an approach based on RGB
camera are summarized in table \ref{tab:results} where the \textit{expected} is
the focus of attention that given the state of the system, it is likely to
attract the subject's gaze. In the same manner, the \textit{captured} is the
focus of attention our approach provides and the \textit{real}, the ground truth
of the subject's gaze direction based on manual annotations from two evaluators
with an average reliability of $ \alpha = .89 $ computed using Cronbach's alpha.

In order to calculate the inter-rater reliability between the two measurements
(real-captured and real-expected), the overlap percentage between the two
curves\fixme{which curve?}
has been computed. A more reliable measurement in terms of time shifting between
the real and captured focuses of attention is provided using the cosine
similarity\fixme{this needs to be better justified: why not a Cronbach alpha?} also shown in table~\ref{tab:results}.

\begin{table*}[h!]
    \centering
    \caption{\textbf{Results of attention tracking accuracy}. \emph{Head pose
    tracking} is the percentage of total time of successful detection of the
    head pose; \emph{Ground-truth/robot match} is the percentage of matching time
    between manually annotated focus of attention (ground-truth) and robot's
    computed focus of attention.}

@@

The results has proved that the VFoA prediction accuracy in the tracking is not
only conditioned by the subject's amount of movement, but also by the speed of
the same movements: Restless children externalize greater levels of movement
decreasing the VFoA prediction accuracy.

For instance, subjects 1,4 and 5 despite the fact that the tracking accuracy is
in the average, the bad precision in the captured VFoA suggests that such
children are highly likely to modify the gaze direction without head movement
intervention reducing dramatically the overlap between the manually annotated 
focus of attention and robot's computed focus of attention. And in consequence,
increasing the standard deviation.

Additionally, several recurrent computer vision challenges such as the lighting
conditions, occlusions with the hands and the face angle with respect to the
camera, looking to the experimenter for instance, are factors that contribute to
reduce the tracking accuracy.

\subsubsection{Subject's Focus of Attention Analysis}

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{matches}
    \caption{\small \textbf{Comparison of computed focus of attention \vs ground
        truth} during a 20+ min face-to-face child-robot interaction (subject 4
        in table~\ref{tab:results}).
        In blue (top line), the focus of attention as computed by the robot;
        in orange (bottom line), the focus of attention as manually annotated
        (ground truth). The bottom graph shows agreements.}

@@

According to figure \ref{fig:realExpected} several findings can be analyzed in
detail. The matching curve (cyan color) shows the overlapping between the
expected and the real focuses of attention. As can be noticed there is a
fluctuation at the beginning of the interaction as well as just before the
activity switching. These chaotic gaze manifestation is what we call an
automatic discovery of events, when the user experiences an uncertainty in the
present situation.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\columnwidth]{bitmap}
    \caption{\small Cosine similarity matrix between ground-truth and robot-measured VFoA
    using a 10 sec. window size. The correspondent numbers are shown
    in figure \ref{fig:realCaptured}.}
    \label{fig:bitmap}
\end{figure}

@@

@@

\begin{table}[h!]
    \centering
    \caption{\textbf{Levels of with-me-ness}. For each subject, the with-me-ness level is reported over the whole interaction, either based on the ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of attention measured by the robot.}
    \begin{tabular}{p{1cm}cccccccc}
        \toprule
        Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg. & SD \\
        \midrule
        $\mathcal{W}_{g.truth}$ & 92.6 & 88.8  & 95.8 & 93.1 & 96.5 & 89.6 & 92.7 & 3.1 \\ 
        %{\it Cronbach's $\alpha$} &  & & & & & & & \\
        %\midrule
        $\mathcal{W}_{robot}$ &  & & & & & & & \\
        %{\it Cronbach's $\alpha$} &  & & & & & & & \\
        \bottomrule
    \end{tabular}
    \label{tab:results-with-me-ness}
\end{table}

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{with-me-ness}
    \caption{\small \textbf{With-me-ness}. Evolution of the level of
    \emph{with-me-ness} over a 20min long child-robot interaction. The diagram
    is identical to Figure~\ref{fig:realExpected} with the \emph{expected}
    attentional targets added in green. The bottom curves represent the levels
    of with-me-ness (since the beginning of the interaction), computed following
    Algorithm~\ref{alg:with-me-ness}. The blue line is with-me-ness as estimated by
    the robot, the orange line is with-me-ness computed from manually annotated
    focuses of attention. The ``holes'' in the with-me-ness computed by the
    robot correspond to period of time where the child's face is not detected.}

@@

If we consider the case shown in section \ref{fromFaceTo}, evaluating the
subject's progression together with the time on task may be an indicator of
engagement: If the subject answers so quickly with respect to the previous
attempts with a poorer handwriting may be the case she is not engaged anymore.
In such hypothetical case an activity switch can be easily implemented in a
robot architecture. It confirms the findings reported in \cite{anzalone} where
time response to event is proposed to account quality of interaction.

>>>>

\section{Experimental Validation}
\label{sec:expe}

@@

Six children (ages 5 to 6, 3 boys, 3 girls) were enrolled for this study.
The study took place at school, in a separate room (the computer
lab). The participants were chosen by the teacher, and would come one after the
other for a face-to-face interaction with a {\sc Nao} robot (average duration:
19.6 min, $SD=1.58$).

The experiment is organized in rounds of writing: during a typical round, the
child requests the robot to write something (a single letter, a number or a full
word), and pushes a tactile tablet (equipped with a custom application that we
developed) towards the robot; the robot writes on the tablet by mimicking the
writing gesture (but without actually physically touching the tablet); the child
then pull back the tablet, corrects the robot's attempt by writing him/herself
on top or next to the robot's writing, and ``send'' his/her demonstration to the
robot by pressing a small button on the tablet. The robot learns from this
demonstration and tries again. The child continues the turn-taking until he/she
decides to train the robot on another word. In total, the children performed in
average 12.16 ($SD=2.61$) rounds of writing.

@@

The experiment was carried with an Aldebaran's {\sc Nao} robot, using ROS as a
middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
Head pose estimation, presented in section~\ref{sec:vfoa}, builds on the {\tt
dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
tf} library. The implementation is open-source and available at
\url{https://github.com/chili-epfl/attention-tracker}.

The implementation of the hand-writing activity itself has been presented by
Hood~\etal in~\cite{Hood:2015}.

\subsection{Data Collection \& Analysis}

Successful detection of the head, and, when detected, the attentional target of
the children as estimated by the robot, were logged during the experiment (in
total, $6\times19.6=117.6$ min of interaction).\fixme{What it the case that 2
focuses of attention were sometimes detected? what did we do in such a
case?}

Besides, we video-recorded the interactions, and performed a {\it post-hoc}
manual coding of the focus of attention by two annotators (Cronbach's
$\alpha=.89$)\fixme{did the 2nd annotator annotated the 6 whole interactions?}.
The manual coding forms our attentional \emph{ground-truth}.

To assess the accuracy of the attention estimation by the robot, we computed
over time (excluding the periods when the head is not detected) the
overlap between the ground-truth and the robot's estimate, the
inter-judge agreement (Cronbach's $\alpha$) and the cosine similarity
matrix.

The cosine similarity is a more reliable measurement between the ground-truth
and the robot measure of the focus of attention as it accounts for temporal
delays.\fixme{this needs to be better justified/explained.}.

@@

The main results are reported in Table~\ref{tab:results} and
Figure~\ref{fig:cosinematrix} (cosine similarity matrix);
Figure~\ref{fig:realExpected} further gives a concrete picture of the
ground-truth \vs computed attentional targets for subject 4 (the subject with
the least successful tracking).

\begin{table*}[h!]
    \centering
    \caption{\textbf{Attention tracking accuracy}. \emph{Head pose
    tracking} is the percentage of total time of successful detection of the
    head pose; \emph{Ground-truth/robot match} is the percentage of matching time
    between manually annotated focus of attention (ground-truth) and robot's
    computed focus of attention. Total duration: 117.6 min.}

@@

\begin{figure}[th!]
    \centering
    \includegraphics[width=0.7\columnwidth]{bitmap}
    \caption{\small Cosine similarity matrix between ground-truth and robot-measured VFoA
    using a 10 sec. window size. The correspondent numbers are shown
    in figure \ref{fig:realCaptured}.}
    \label{fig:cosinematrix}
\end{figure}

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{matches-excerpt}
    \caption{\small \textbf{Comparison of computed focus of attention \vs ground
        truth} during a face-to-face child-robot interaction (subject 4
        in table~\ref{tab:results}, 3.5min-long excerpt).
        In blue (top line), the focus of attention as computed by the robot;
        in orange (bottom line), the focus of attention as manually annotated
        (ground-truth). The bottom graph shows agreement between both (whenever
        the head is detected).}

@@

\subsubsection{Relevance of Head Pose Estimation for Attention Tracking}
\label{sec:fromFaceTo}

During the whole interaction, the head pose of the children was satisfactorily
and consistently tracked (86\% of the time in average, $SD=3.0$), which is
expected for a face-to-face interaction with a static head-tracking camera
(meaning that the child head would remain in the field of view of the camera
most of the time). This is however comforting in terms of suitability of our
approach for head pose estimation with children in field experiments.

In our experiment, the primary causes of lost head pose were occlusions with the
hands (similar to the middle-bottom picture in
Figure~\ref{head_pose_real_world}), close proximity with the tablet while
writing, and gaze directed to the facilitator (sitting directly on the left of
the child, Figure~\ref{fig:realSetup}).

In terms of attention tracking, two groups of children stand out of
Table~\ref{tab:results}: the robot's estimate of the focus of attention for
subjects 1, 4 and 5 has questionable agreement with the ground-truth (Cronbach's
$\alpha$ close to 0.6), while for subjects 2, 3 and 6, we find a good agreement
($\alpha$ close to 0.8). This does not correlate with the success rate of the
head tracking itself, and, after further analysis of the videos, it rather
reflects that the first group of children rely more on their eyes to direct
their gaze than the second group. However, and notwithstanding the small sample
size, it is rather intriguing that the two groups appear so clearly separated.
This certainly calls for further experimental investigation.

@@

\subsection{Concept \& Calculation}

@@

\fixme{mention somewhere probabilistic extensions of withmeness}

\subsection{Experimental Interpretation}

During the experiment presented above (section~\ref{sec:expe}), the robot
controller associated a (set of) expected attentional targets to the task being
performed. For instance, while the robot was waiting for the child's handwriting
demonstration, the expected attentional target was the tablet (since the child
was supposed to write there), and while the robot was writing itself, the
attentional targets we expected to be either on the tablet or on the robot.
These expected targets, forming together the robot's attentional {\it a priori},
are depicted on Figure~\ref{fig:with-me-ness} as the green lines.

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{with-me-ness}
    \caption{\small \textbf{With-me-ness}. Evolution of the level of
        \emph{with-me-ness} over the whole $\approx$20min long interaction of
        subject 4. \emph{Interaction Episodes} describe the distinct phases of
        the interaction, and are discussed in the text. The central chart is
        similar to Figure~\ref{fig:realExpected} with the \emph{expected}
        attentional targets added in green. The bottom diagram represents the
        levels of with-me-ness over successive windows of 30 seconds. The blue
        line is with-me-ness as estimated by the robot, the orange line is
        with-me-ness computed from manually annotated attentional targets. The
        ``holes'' in the with-me-ness computed by the robot correspond to period of time
        where the child's face is not detected.}

@@

\begin{table}[h!]
    \centering
    \caption{\textbf{Levels of with-me-ness}. For each subject, the with-me-ness level is reported over the whole interaction, either based on the ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of attention measured by the robot.}
    \begin{tabular}{p{1cm}cccccccc}
        \toprule
        Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg. & SD \\
        \midrule
        $\mathcal{W}_{g.truth}$ & 92.6 & 88.8  & 95.8 & 93.1 & 96.5 & 89.6 & 92.7 & 3.1 \\ 
        $\mathcal{W}_{robot}$ &  & & & & & & & \\
        \bottomrule
    \end{tabular}
    \label{tab:results-with-me-ness}
\end{table}

According to figure \ref{fig:with-me-ness} several findings can be analyzed in
detail. The matching curve (cyan color) shows the overlapping between the
expected and the real focuses of attention. As can be noticed there is a
fluctuation at the beginning of the interaction as well as just before the
activity switching. These chaotic gaze manifestation is what we call an
automatic discovery of events, when the user experiences an uncertainty in the
present situation.

If we consider the case shown in section \ref{sec:fromFaceTo}, evaluating the
subject's progression together with the time on task may be an indicator of
engagement: If the subject answers so quickly with respect to the previous
attempts with a poorer handwriting may be the case she is not engaged anymore.
In such hypothetical case an activity switch can be easily implemented in a
robot architecture. It confirms the findings reported in \cite{anzalone} where
time response to event is proposed to account quality of interaction.
